\addchap{Appendix 1: Sample text, Ashret dialect}

\section*{Across the Lowari Pass}

This is an account given by Muhammad Hussain, son of Noor Majid, recorded on 23 July 2003 by Naseem Haider, in the former’s home in \textit{Čhíni} in the upper Ashret Valley. Muhammad Hussain, father of four sons and three daughters, and the late elder of the \textit{Kučurooṭeé} clan, was born in 1920 and passed away in 2007. He served as a postman during the time of the \textit{ṣóo}, carrying letters across the Lowari Pass between Chitral and Dir. Muhammad Hussain was not formally educated but spoke Pashto and Khowar. The narrative is a recollection of a dramatic event taking place at the Lowari Pass. While carrying the prime minister’s wife in a sealed carriage through the pass, Muhammad Hussain and his comrades were swept away by a sudden avalanche but managed to get out of the snow and come to the rescue of the prominent lady.


\begin{exe}
\ex
\label{ex:A1-1}
\glll asaám	the	húkum	bhílu	hukumatíi	ki waziirazamíi	yaaní	bibí	yhéendi \\
asaám	the	húkum	bhe-íl-u	hukumát-ii	ki waziirazám-ii	yaaní	bibí	yhéi-ánd-i \\
\textsc{1pl.acc} to	order	become-\textsc{pfv"=msg} government-\textsc{gn} \textsc{comp} prime.minister-\textsc{gn} that.is	wife	come-\textsc{prs"=f} \\
\glt We were told by the government that the wife of the Prime Minster would be coming,

\ex
\label{ex:2}
\glll tus	kháaṇa	the	bóoi \\
tus	kháaṇ-a the	be-ooi \\
\textsc{2pl.nom} mountain-\textsc{ob} to go-\textsc{imp.pl} \\
\glt and that we should therefore head to the Lowari Pass.

\ex
\label{ex:3}
\glll be	wáxta	be	táa	be	bhéṭa \\
be	wáxta	be-í	táa	be-í	bhéṭ-a \\
\textsc{1pl.nom}	early	go-\textsc{cv}	there.\textsc{rem}	go-\textsc{cv}	sit.\textsc{pfv"=mpl} \\
\glll bhéṭa	bhéṭa	bhéṭa \\
bhéṭ-a	bhéṭ-a	bhéṭ-a \\
sit.\textsc{pfv"=mpl}	sit.\textsc{pfv"=mpl}	sit.\textsc{pfv"=mpl} \\
\glt Early in the morning we went there and waited for a long time.

\ex
\label{ex:4}
\glll yaaní	ḍoolái	darák	na	bhíli \\
yaaní	ḍoolái	darák	na	bhe-íl-i \\
that.is	carriage	trace	\textsc{neg}	become-\textsc{pfv"=f} \\
\glt There was no news about the carriage.

\ex
\label{ex:5}
\glll maaxaám	c̣hiṇ	bhílu \\
maaxám	c̣hiṇ	bhe-íl-u \\
evening	darkness	become-\textsc{pfv"=msg} \\
\glt In the evening, it became dark.

\ex
\label{ex:6}
\glll the	ɡhueeṇíia	ɡhiní	phedíla \\
the-í	ɡhueeṇíi-a	ɡhín-í	phed-íl-a \\
do-\textsc{cv}	Pashtun-\textsc{pl}	take-\textsc{cv}	arrive-\textsc{pfv"=mpl} \\
\glt Then, finally, some Pashtuns arrived with it.

\ex
\label{ex:7}
\glll phedí	asaám	the	hawaalá	thíilu \\
phed-í	asaám	the	hawaalá	the-íl-u \\
arrive\textsc{cv}	\textsc{1pl.acc}	to	custody	do-\textsc{pfv"=msg} \\
\glt They handed it over to us.

\ex
\label{ex:8}
\glll máathe	míi	nóo	de	maníitu	ki	asíi \\
ma=the	míi	náau	de-í	maníit-u	ki	asíi \\
\textsc{1sg.nom}=to \textsc{1sg.gn}	name	give-\textsc{cv}	say.\textsc{pfv"=msg}	   \textsc{comp} \textsc{1pl.gn} \\
\glll ráaǰamii	zimawaár	tu \\
raǰ-am-ii	zimawaár	tu \\
rope-\textsc{pl.ob"=gn}	responsible	\textsc{2sg.nom} \\
\glt They called my name and told me to take responsibility for their ropes.

\ex
\label{ex:9}
\glll be	ḍoolái	na	mučáa	bhaáana \\
be	ḍoolái	na	mučá-áa	bha-án-a \\
\textsc{1pl.nom}	carriage \textsc{neg}	open-\textsc{inf}	be.able.to-\textsc{prs"=mpl} \\
\glt We cannot undo [the ropes of] the carriage.

\ex
\label{ex:10}
\glll míi	maníitu	ki	šóo	tus	bóoi \\
míi	maníit-u	ki	šu-u	tus	be-ooi \\
\textsc{1sg.erg}	say.\textsc{pfv"=msg} \textsc{comp}	good-\textsc{msg} \textsc{2pl.nom}	go-\textsc{imp.pl} \\
\glt I said, “That’s fine, you may leave!”

\ex
\label{ex:11}
\glll be	oóra	xoneeṛí	thaní	áa	néeṛi	híni \\
be-í	oóra	xoneeṛí	thaní	a	néeṛi	hín-i \\
go-\textsc{cv}	over.here	Khoneri	\textsc{qt}	\textsc{idef}	brook	be.\textsc{prs"=f} \\
\glll xatarnaák	néeṛi,	dhrúuk \\
xatarnaák	néeṛi	dhrúuk \\
dangerous	brook	gorge \\
\glt There is on this side a brook, called Khoneri, that is a dangerous brook, or gorge.

\ex
\label{ex:12}
\glll be	eetáa	yhóola	ta	hiimaál	čhinǰí asaám	híṛa \\
be	eetáa	yhéi-íl-a	ta	hiimaál	čhinǰ-í asaám	híṛ-a \\
\textsc{1pl.nom}	there.\textsc{rem}	come-\textsc{pfv"=mpl}	\textsc{ds}	glacier	strike-\textsc{cv} \textsc{1pl.acc}	take.away.\textsc{pfv"=mpl} \\
\glt When we reached there, an avalanche hit and swept us away,

\ex
\label{ex:13}
\glll bhun	áa	lhaáṣṭ	zhayí	de \\
bhun	a	lhaáṣṭ	zhaáy-í	de \\
down.below	\textsc{idef}	plain	place-\textsc{ob}	\textsc{pst} \\
\glt to a flat place below.

\ex
\label{ex:14}
\glll se	lhaṣṭíwee	de	asaám	be \\
se	lhaáṣṭ-í=wée	de-í	asaám	be \\
\textsc{def}	plain-\textsc{ob}=into	give-\textsc{cv} \textsc{1pl.acc} \textsc{1pl.nom} \\
\glll dharíita \\
dharíit-a \\
remain.\textsc{pfv"=mpl} \\
\glt We were carried away to that flat place and came to rest there.

\ex
\label{ex:15}
\glll dharí	dúi	hiimaál	whaí	ṭópa	traác̣  \\
dhar-í	dúi	hiimaál	whéi-í	ṭópa	traác̣  \\
remain\textsc{cv}	another	glacier	come.down-\textsc{cv}	down \textsc{host} \\
\glll de	asaám	rhalá	ɡaḍíla \\
de-í	asaám	rhalá	ɡaḍé-íl-a \\
give-\textsc{cv} \textsc{1pl.acc}	up	pull.out-\textsc{pfv"=mpl} \\
\glt Then another avalanche came and brought us out.

\ex
\label{ex:16}
\glll rhalá	ɡaḍílii	baád	alaahirahúm \\
rhalá	ɡaḍé-íl-ii	baád	alaahirahúm \\
up	pull.out-\textsc{pfv"=gn}	after	by.grace.of.God \\
\glt After coming out, by the mercy of God,
 
\ex
\label{ex:17}
\glll be	khilaí	ba	na	de \\
be	khilaí	ba	na	de \\
\textsc{1pl.nom}	alone \textsc{top} \textsc{neg} 	\textsc{pst} \\
\glt we were not alone.

\ex
\label{ex:18}
\glll yaaní	iskoóṭ	de	abdulxaalíq	de \\
yaaní	iskoóṭ	de	abdulxaalíq	de \\
that.is	scout \textsc{pst}	<person>	\textsc{pst} \\
\glt A scout and Abdul Khaliq were with us,

\ex
\label{ex:19}
\glll yaaní	iskooṭíi	puluswaalá	de \\
yaaní	iskoóṭ-ii	puluswaalá	de \\
that.is	scout-\textsc{gn}	police.man \textsc{pst} \\
\glt and so were some scouts and police.

\ex
\label{ex:20}
\glll buṭheé	the	asaám	híṛa \\
buṭheé	the-í	asaám	híṛ-a \\
all	do-\textsc{cv}	\textsc{1pl.acc}	take.away.\textsc{pfv"=mpl} \\
\glt All of us had been swept away.

\ex
\label{ex:21}
\glll rhalá	nikháatii	baád	asím	dhuím	ǰáanam \\
rhalá	nikhát-ii	baád	asím	dhuím	ǰáan-am \\
up	come.out.\textsc{pfv"=gn}	after	\textsc{1pl.erg}	two.\textsc{ob}	person-\textsc{pl.ob} \\
\glll tas	se	hiimeelíi	díi	dhrakí	ɡaḍíli \\
tas	se	hiimaál-ii	díi	dhraké-í	ɡaḍé-íl-i \\
\textsc{3sg.acc} \textsc{def}	glacier-\textsc{gn}	from	pull-\textsc{cv}	pull.out-\textsc{pfv"=f} \\
\glt after getting ourselves out, two of us pulled her out of the glacier as well,

\ex
\label{ex:22}
\glll áa	ta	ma	mhaatuseén	míi	nóo \\
a	ta	ma	mhaatuseén	míi	náau \\
one	\textsc{ds} \textsc{1sg.nom}	<person> \textsc{1sg.gn}	name \\
\glt One of them was me, my name is Muhammad Hussain,

\ex
\label{ex:23}
\glll áa	ba	habibulaxaán	thaní	míiš	de \\
a	ba	habibulaxaán	thaní	méeš	de \\
one \textsc{top}	<person>	\textsc{qt}	man	\textsc{pst} \\
\glt and one was a man named Habibullah Khan,

\ex
\label{ex:24}
\glll marí	hínu	xudaái	tas	ubaxíi \\
mar-í	hín-u	xudaái	tas	ubaxé-e \\
die-\textsc{cv}	be.\textsc{prs"=msg}	God	\textsc{3sg.rem.acc}	forgive-\textsc{3sg} \\
\glt who is dead now, may God grant him forgiveness.

\ex
\label{ex:25}
\glll ɡaḍílii	baád	yaaní	tíi	asíi	nóo \\
ɡaḍé-íl-ii	baád	yaaní	tíi	asíi	náau \\
pull.out-\textsc{pfv"=gn}	after	that.is	\textsc{3sg.rem.ob}	\textsc{1pl.gn}	name \\
\glll čooṇṭaá \\
čooṇṭá-í \\
write-\textsc{cv} \\
\glt After we got her out she wrote down our names,

\ex
\label{ex:26}
\glll yaaní	asaám	inaám	bi	bíiḍi	díti \\
yaaní	asaám	inaám	bi	bíiḍ-i	dít-i \\
that.is	\textsc{1pl.acc}	gift \textsc{sep}	much-\textsc{f}	give.\textsc{pfv"=f} \\
\glt and gave us many gifts.

\ex
\label{ex:27}
\glll yaaní	míi	zueeníi	daureení	eeṛé	haalaát \\
yaani	míi	zuaán-ii	dauraán-í	eeṛé	haalaát \\
that.is	\textsc{1sg.gn}	youth-\textsc{gn}	period-\textsc{ob}	\textsc{dist}	condition \\
\glll ma	baándi	yhéeli \\
ma	baándi	yhéi-íl-i \\
\textsc{1sg.nom}	on	come-\textsc{pfv"=f} \\
\glt This event happened to me during my youth.

\end{exe}